Parallel Acceleration of Fixed Complexity Sphere Decoder for Large MIMO Uplink System based on GPU

### 2014-10-28 parallel acceleration of FCSD based on cuda/draft.tex
revised the latex draft

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
The large MIMO system which is a scaling up of conventional MIMO systems with several hundreds of antennas at the base station serving several tens of terminals at the same time,  which is thought to be a big idea for the next generation wireless communication system. However the detection of the large MIMO system acquires a long time to get optimum or sub-optimum solutions, In order to accelerate the real-time decoder for large MIMO system, in this paper we present General Purpose Graphic Processing Units (GPGPU) implementation of Fixed Complexity Sphere Decoder (FCSD) for the large scale MIMO system based on the dominant propriety framework of GPGPU-Computed Unified Device Architecture (CUDA). The simulation shows a up to nearly $5$ times acceleration for $32\times 32$ array of whole application CPU version, furthermore, the proposed CUDA-FCSD algorithm shows a increasing speedup performance for a increasing array size. 
\end{abstract}

@@

 The common challenge that all the researchers, engineers and industrial companies are now facing in wireless communication area is the sharp contrast between the increasing demand of the mobile network speed for the transmission of rich media content and better QoS (quality of service) and the limited bandwidth of the radio frequency spectrum and a complex space time varying wireless environment, moreover in foreseeable future this demand will continue grow. Multiple Input Multiple Output (MIMO) has become a very promising technology that can help to solve this problem, the core idea of MIMO is to use multiple antennas both at the transmitting end and receiving end, the spatial domain sampled signal can be combined to provide parallel data pipes so that it can provide array gain, spatial diversity gain, spatial multiplexing gain, and interference reduction\cite{oestges2010mimo}. LTE standard support up to 8 antennas at the base station\cite{dahlman20103g}, The large scale MIMO (also named massive MIMO) is a system with hundreds of low-power low-price antennas collocated at the base station and serving several tens of terminals at the same time\cite{rusek2013scaling}, originated from multiuser MIMO (MU-MIMO)\cite{larsson2013massive}. Large scale MIMO will inherit and enhance all the advantages of conventional MIMO (array gain, spatial diversity gain, spatial multiplexing gain and interference reduction), as well as reduction of latency, improvement of the whole system robustness. 
 As a classical list based decoder for MIMO system, Sphere decoder (SD) can reach the optimum diversity gain as maximum likelihood (ML) detection with a dramatic complexity reduction\cite{viterbo1999universal}\cite{hassibi2005sphere}, however, there are two major problems of this decoder is the varied complexity with signal to noise ratio and sequential nature, which is contradictory with the requirement of a constant data processing throughput and parallel implementation for practical system. Fixed Complexity Sphere Decoder (FCSD), overcomes this two disadvantage of the SD by searching the solution along the fixed number independent multiple paths\cite{barbero2008fixing}.
GPGPU (General purpose Graphic processing Units Computing) has become a big idea in the industry and research of a variety of areas that needs massive data analysis and computation such as computer vision\cite{fung2008using}, computer games\cite{blewitt2013applicability}, signal processing\cite{van2011accelerating}\cite{6671435} and finance\cite{grauer2013accelerating}. Comparing to CPU, The GPU has been designed as computational intensive, with most of the hardware resource be allocated for data processing and parallel computation rather than caching and flow control. In recent years GPU has reached huge advantages in data processing speed over CPU\cite{nvidia2008programming}.  CUDA (Computed unified Device Architecture) is a programming model for GPGPU that proposed by Nvidia which has becoming the dominant proprietary framework for GPGPU programming, it is based on popular programming languages(C, C++, java, python, Fortran, Direct Compute) so that the application developers in different areas do not need to learn a new programming language or get to know the hardware details. The core essence of the CUDA is the Single Instruction Multiple Threads (SIMT) programming model, enables GPU can accelerate a lot of applications with the large dataset by implementing parallel data processing of complex computations. CUDA has become a parallel computing platform for wide community of different application developers in different areas.  

@@

\section{CUDA GPU Architecture and  Programming Model} \label{programming model}
  CUDA programming model is the C programming extension, with special defined syntax and structure, CUDA divide the CPU and GPU as host and device, the serial code executes on the host (CPU) side and parallel data processing is performed on the device (GPU) side, which is defined as heterogeneous programming model. The code executes on the device side is defined as $\mathit{kernel}$ function, which is a single instruction invoked at host side but with a massive parallel data processing threads execution on the device (SIMT), This programming architecture is askin to Single Instruction Multiple Data (SIMD), however unlike the SIMD vectoring machines, SIMT can branching the behavior of each threads so that reach thread-level parallelism, thread is the basic programming unit of CUDA, a block of cooperative threads form a thread block, in one block all the threads can share a memory called shared memory that all the threads in the block have the access to this pitch of memory, the GPU use massive number of threads to hide the long latency of global memory instead of using large cache as CPU, Another function of the thread block is all the thread execute in one block can synchronize their action. The warp is the basic schedule unit with a constant number of threads  (for current version of CUDA device, the warp size is 32), when one warp are waiting for data loading the other warps just begin to execute. In a word CPU cores are designed to minimize latency for each thread at a time , whereas GPUs are designed to handle a large number of concurrent, lightweight threads in order to maximize throughput. Blocks are grouped into a one-dimensional, two-dimensional, or three-dimensional grid, similarly, one thread block can also be organized into one-dimentional, two-dimensional or three-dimensional. The organization of the block and grid shape and dimension is determined by the concrete application, Figure \ref{figure1} shows the 3 dimensional thread hierarchy of CUDA.
\includegraphics[scale=0.5]{CUDA_programming_hierachy.eps}
\caption{CUDA 3D thread hierachy\\ $5\times 5\times 5$ grid and each block contains $5\times 5\times 5$ threads\\}
\label{figure1}
In Figure \ref{figure2} we use the GPU-GeForce GTX 760 as an example to show the microarchitecture of  GPU. The GPU works as a coprocessor system, consists of Stream Multiprocessors (SM), GeForce GTX 760 has 6 SM performs the multiply-add arithmetic operation and furthermore, each SM has the special functional units(SFU) that can perform more complex arithmetic operation such as trigonometric functions and reciprocal square root. the shared memory is also called on-chip memory which is located at SM that has low latency and limited capacity. Global memory, which is off-chip located memory of GPU with long latency and large capacity, another two types of the memory is called texture memory and constant memory, which are located off-chip , but cached, therefore the loading speed of the data located at these two type of memory space will be much faster than the that from global memory, for constant memory, it is read only and the memory latency is approximately as small as the registers.  The variables invoked in kernel functions can be stored in either of the these 4 types of memories. Registers are allocated dynamically to threads.\\
   Since the resource limitation of SM the total number of threads that we can define in one block is limited, on current CUDA supported GPUs, one block can contains up to 1024 threads, in addition all the blocks are executed on SMs, thus the total number of threads we can make use of is $T_{total}=Num_{SM}\times Num_{threads/SM}$, GeForce GTX 760 allows 2048 threads in one SM, thus the total number of threads that can work in parallel is $2048\times 6=12288$, although in concrete application, the programmer can allocate more than 12288 threads for a kernel, but the threads will work in serial, which will influence the performance of program. 

@@

\section{GPU Based Acceleration of FCSD}

@@

\section{Simulation Results}

@@

\section{Conclusion}

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Large MIMO constitutes a principal techniques for next generation wireless communication systems. Detection algorithms for large MIMO tend to require long simulation times and on conventional computer systems, In order to reduce the simulation times for decoders of large MIMO systems. In this paper we propose to use General Purpose Graphic Processing Units (GPGPU)to implement Fixed Complexity Sphere Decoder (FCSD) for the large scale MIMO system based on the dominant framework of GPGPU-Computed Unified Device Architecture (CUDA). The proposed CUDA-FCSD can chieve up to five times speedup over conventional CPU implementation while reaching the same Bit Error Rate (BER) performance. Furthermore the proposed algorithm demonstrates great potential to accelerate large MIMO systems.
\end{abstract}

@@

 \paragraph{} The present challenge in the wireless communication area is the sharp contrast between the increasing demand of mobile network speed for the transmission of rich media content with better QoS (quality of service), and the shortage of radio frequency spectrum. Multiple Input Multiple Output (MIMO) is thought to be a very promising technology that can help to solve this problem owing to several advantages it can offer. The core idea of MIMO is to use multiple antennas that at the transmitting and receiving sides that provide parallel data pipes in space resulting in spatial multiplexing gains\cite{oestges2010mimo}. For example the LTE standard supports up to 8 antennas at the base station\cite{dahlman20103g}. 
 \paragraph{} Large scale MIMO (also named massive MIMO) systems employ hundreds of low-power low-price antennas collocated at the base station and serving several tens of terminals at the same time\cite{rusek2013scaling}. This technology originated from multiuser MIMO (MU-MIMO) systems\cite{larsson2013massive}. Large scale MIMO will inherit and enhance all the advantages of conventional MIMO (spatial diversity gain and spatial multiplexing gain), as well as reduction of latency and improvement of system robustness. 
 A common decoder for MIMO system is the Sphere Decoder (SD) that perform as well as a Maximum Likelihood (ML) decoder with a dramatic complexity reduction\cite{viterbo1999universal}\cite{hassibi2005sphere}. There are two major problems with the traditional SD decoder: complexity depending on Signal to Noise Ratio (SNR), and sequential nature, which is contradictory with the requirement of a constant data processing throughput and parallel implementation for practical system. The Fixed Complexity Sphere Decoder (FCSD) overcomes these two disadvantages of the SD decoder by searching the solution along a fixed number of independent paths\cite{barbero2008fixing}.
\paragraph{}General purpose Graphic processing Units Computing (GPGPU) has become a common trend in industry and research in areas that need massive computations such as computer vision\cite{fung2008using}, computer games\cite{blewitt2013applicability}, signal processing\cite{van2011accelerating}\cite{6671435} and finance\cite{grauer2013accelerating}. Compared to CPU, the GPU can naturally cope with computational intensive tasks, since most of the hardware resource are allocated for data processing and parallel computation, rather than caching and flow control. In recent years GPU computing has provided significant advantages in data processing speeds over CPU computations\cite{nvidia2008programming}. Computed unified Device Architecture (CUDA) is a programming model for GPGPU that was proposed by Nvidia and is becoming the dominant framework for GPGPU programming. It is based on popular programming languages(C, C++, Java, Python, Fortran, Direct Compute), such that the application developers in different areas do not need to learn a new programming language or get to know the hardware details. CUDA employs the Single Instruction Multiple Threads (SIMT) programming model, enables a GPU implement parallel data processing, and has become a parallel computing platform for wide community of different applications.
\paragraph{}In this paper, a GPU based implementation of Fixed Complexity Sphere Decoder is proposed, A heterogeneous programming solution for FCSD preprocessing and blocked parallel path searching solution are presented, the speedup results of different influence factors are presented and analyzed, The proposed CUDA-FCSD implementation demonstrates a great potential for large MIMO systems.
\paragraph{}The rest of the paper is organized as follows. Section \ref{system} describes the MIMO systems model. Section \ref{programming model} introduces CUDA GPU architecture and the programming model. In section \ref{GPUFCSD}, we present details of the implementation of CUDA-FCSD. In section \ref{simulation}, the speedup and bit error rate (BER) performance of CUDA-FCSD are presented and analyzed. Finally conclusions are drawn in section \ref{conclusion}.    

@@

\section{CUDA GPU Architecture and  Programming Model} \label{programming model}
\subsection{Memory Hierarchy}  
CUDA programming model is the C programming extension, with special defined syntax and structure. The serial code executes on the host (CPU) side, and massive parallel data processing is performed on the device (GPU) side, is defined as a heterogeneous programming model. A thread is the basic programming unit of CUDA, a block of cooperative threads form a thread block, in one block all the threads can share one pitch of on chip memory (shared memory), all the threads executed in one block can synchronize their action meaning the threads are permitted to run to the next instruction only when all the threads have reached the same instruction node. Blocks can be grouped into a one-dimensional, two-dimensional, or three-dimensional grid. Similarly, The threads can also be organized into one-dimensional, two-dimensional or three-dimensional blocks. The organization of the block and grid shape and dimension is determined by the application, Figure \ref{figure1} illustrates the 3 dimensional thread hierarchy of CUDA.
\begin{figure}[htb]
\centering
\includegraphics[scale=0.5]{CUDA_programming_hierachy.eps}
\caption{CUDA 3D thread hierachy\\ $5\times 5\times 5$ grid and each block contains $5\times 5\times 5$ threads\\}
\label{figure1}
\end{figure}

\subsection{Single Instruction Multiple Thread (SIMT) Model} Threads are scheduled, managed, created and executed in a group of 32 threads called warp, the programs executes on the device side are defined as $\mathit{kernel}$ function, when a $\mathit{kernel}$ function is invoked, the thread blocks are partitioned into warps, the threads in one warps executes the same instruction simultaneously, the different warps are scheduled by warp scheduler, when one warp is waiting for data loading the other warps begin to execute, this is called Single Instruction Multiple Thread (SIMT). This programming architecture is askin to Single Instruction Multiple Data (SIMD), however unlike the SIMD vectoring machines, SIMT can branch the behavior of each threads to achieve thread-level parallelism. The GPU use massive number of threads to hide the long latency of global memory instead of using large cache as CPU, in a word CPU cores are designed to minimize latency for each thread at a time, whereas GPUs are designed to handle a large number of concurrent, lightweight threads in order to maximize throughput.
\subsection{Mircroarchitecture of GPU} In Figure \ref{figure2} we use the GPU-GeForce GTX 760 as an example to show the microarchitecture of  GPU. The GPU works as a coprocessor system, consisting of Stream Multiprocessors (SM). The GeForce GTX 760 has 6 SM that perform the multiply-add arithmetic operation. Furthermore, each SM has the special functional units(SFU) that can perform more complex arithmetics such as trigonometric functions and reciprocal square root. The shared memory is on chip memory located at SM and has low latency and limited capacity. Global memory, which is off-chip memory of the GPU with long latency and large capacity. Another two types of the memory, called texture memory and constant memory, are located off-chip, but cached. Therefore the loading speed of the data located in these two type of memory space is much faster than that from global memory. Constant memory is read only and with latency approximately as small as the registers.  The variables invoked in kernel functions can be stored in either of the these four types of memories. Registers are allocated dynamically to threads.
\paragraph{}Because of the resource limitation of SM the total number of threads can be defined in one block is limited. On current CUDA supported GPUs, one block can contains up to 1024 threads, and all the blocks are executed on SM. Thus the total number of threads ($T_{total}$) we can make use of is $T_{total}=N_{threads}\times N_{threads/SM} $, $N_{SM}$ and $N_{threads/SM}$ denote the number of SM and number of threads per SM. The GeForce GTX 760 allows 2048 threads in one SM, thus the total number of threads that can work in parallel is $2048\times 6=12288$. In concrete applications, the programmer can allocate more than 12288 threads for a kernel, but the threads will work in serial, which will influence the program performance. 

@@

\section{GPU Based Acceleration of FCSD}\label{GPUFCSD}

@@

\section{Simulation Results}\label{simulation}

@@

\section{Conclusion}\label{conclusion}
